# (APPENDIX) Appendix {-} 

# Using *tidycensus* to extract relevant census data

In Section 5.1, we read in a county-level demographic dataset (extracted from the 2010 decennial census), which we subsequently used to generate one of the components of our bias indicator. That dataset was provided to you at the start of the workshop, and we did not discuss the process by which the dataset was generated in the main lesson. However, if you are curious about how that data was generated, the script in this Appendix can be used as a guide to reproduce the demographic dataset introduced in Section 5.1.  

To extract the census demographic data required to create the bias index, we must first load a few required packages. In particular, please load the ```tidyverse``` (which was used in the workshop tutorial) and the ```tidycensus``` package, which is an R package that allows users to pull Census Bureau data using the Census API (if *tidycensus* is not already installed, please install it using ```install.packages("tidycensus")```.

```{r, message=F, warning=F, results=F}
# Loads libraries required to extract census data
library(tidycensus)
library(tidyverse)
```

Before extracting data with ```tidycensus```, you must [acquire a Census API key from the Census Bureau website](https://api.census.gov/data/key_signup.html); once you apply for a key on the website, your key will be immediately emailed to the address you provide. Enter your census API key in R Studio with the following code, replacing the <INSERT HERE> with your key:

```{r, eval=FALSE}
census_api_key("INSERT HERE")
```

In the workshop, we were working with 2010 police stops data, so it made sense to pull demographic data from the 2010 decennial census (which would have been collected in 2009). 
The discussion below is therefore framed with respect to the 2010 decennial census; if you choose to use another census data product (such as the American Community Survey) to create the bias index, your code may look slightly different. 

##  Define your variables

First, we can generate a table that contains the various variables (and associated variable codes) for the 2010 decennial census by using the **```load_variables()```** function. The arguments to the function below ```(2010, "sf1")``` indicate that we want to extract variable names and codes for the 2010 decennial census. Below, we assign the table to an object named ```decennial_2010_variables```: 

```{r}
# Variable list for 2010 Decennial assigned to object named "decennial_2010_variables"
decennial_2010_variables<-load_variables(2010, "sf1")
```

At this point, we can easily view the 2010 decennial variables table by passing ```decennial_2010_variables``` as an argument to the **```View()```** function: 

Based on the information in ```decennial_2010_variables```, we can identify the variable codes for the variables we want to extract. Then, we will define a vector (see below), assigned to an object named ```my_vars``` that assigns the variable codes to descriptive names; these descriptive names will be used as column names in the dataset returned by the census API call, while the variable codes will be used by *tidycensus* to populate the respective fields with the desired data. 

For the purpose of defining the "bias_index" variable, recall that we needed to define the county-level percentage of traffic stops involving a Black motorist, and the county-level percentage of the adult Black population (with respect to the county's overall adult population). The first component of this index was derived using the Stanford Open Policing data. In order to calculate the second component of the bias index, we need two key variables at the county level: the total adult (which we define as 17+) population, and the total adult Black population. These two variables can be extracted from the decennial census. However, there is no separate category in the decennial census for these two measures, so we must derive them based on the data that is available. 

Therefore, given the available data, calculating the total over-17 population will require us to extract data for the male under-5 population, the male 5 to 9 population, the male 10 to 14 population, and the male 15 to 17 population, and analogous measures for the female population. Subtracting these values from the total overall population (among all age groups) will yield a value for the  total over-17 population. 

To calculate the Black over-17 population, we will extract a variable that defines the total Black population, and a series of variables that indicate the Black population for different demographic (sex/age) combinations  17 years old and younger; subtracting the sum of the latter variables from the total Black population yields a value for the Black over-17 population. 

The code below extracts all the variables needed to carry out these calculations:

```{r}
# Define and name variables for census API call
my_vars<-c(total_pop="P001001",
           totalpop_men_u5="P012003",
           totalpop_men_5to9="P012004",
           totalpop_men_10to14="P012005",
           totalpop_men_15to17="P012006",
           totalpop_women_u5="P012027",
           totalpop_women_5to9="P012028",
           totalpop_women_10to14="P012029",
           totalpop_women_15to17="P012030",
           black_totalpop="PCT012B001",
           black_men_u1="PCT012B003",
           black_men_1="PCT012B004",
           black_men_2="PCT012B005",
           black_men_3="PCT012B006",
           black_men_4="PCT012B007",
           black_men_5="PCT012B008",
           black_men_6="PCT012B009",
           black_men_7="PCT012B010",
           black_men_8="PCT012B011",
           black_men_9="PCT012B012",
           black_men_10="PCT012B013",
           black_men_11="PCT012B014",
           black_men_12="PCT012B015",
           black_men_13="PCT012B016",
           black_men_14="PCT012B017",
           black_men_15="PCT012B018",
           black_men_16="PCT012B019",
           black_men_17="PCT012B020",
           black_women_u1="PCT012B107",
           black_women_1="PCT012B108",
           black_women_2="PCT012B109",
           black_women_3="PCT012B110",
           black_women_4="PCT012B111",
           black_women_5="PCT012B112",
           black_women_6="PCT012B113",
           black_women_7="PCT012B114",
           black_women_8="PCT012B115",
           black_women_9="PCT012B116",
           black_women_10="PCT012B117",
           black_women_11="PCT012B118",
           black_women_12="PCT012B119",
           black_women_13="PCT012B120",
           black_women_14="PCT012B121",
           black_women_15="PCT012B122",
           black_women_16="PCT012B123",
           black_women_17="PCT012B124")
```

## Extract the variables using tidycensus

Now that we have a vector of the variables we want to extract (along with descriptive names for those variables), we will use the **```get_decennial()```** function from *tidycensus* to extract these variables from the 2010 decennial census. Several arguments are passed to the **```get_decennial()```** function below:

* ```geography="county"``` specifies that we want the census data to be provided at the county level
* ```variables=my_vars``` specifies the variables we want to extract, and the names they are to be given in the dataset; this information is contained in the ```my_vars``` vector defined above
* ```state=CO``` specifies the state for which we want to extract the data; this argument, together with the ```geography="county"``` argument, means that *tidycensus* will extract the specified data in ```my_vars``` at the county level for the state of Colorado. 
* ```survey="sf1"``` indicates which census dataset we would like to query; here ```sf1``` (short for Summary File 1), indicates we are referring to the decennial census (as opposed, for example, to the American Community Survey)
* ```output=wide``` indicates that we want the dataset with the extracted variables to be in "wide" format, wherein each variable is assigned to its own column. 
* ```year=2010``` indicates that we are interested in data from 2010. Combined with ```survey="sf1"```, this will extract the 2010 decennial census data. 

Finally,  we'll assign the extracted dataset to an object named ```co_counties_race```:

```{r}
# Issue call to Census API
co_counties_race<-
get_decennial(
  geography="county", 
  variables=my_vars,
  state="CO",
  survey="sf1",
  output="wide",
  year=2010)          
```

We can print the first few lines of the dataset to the console to view its structure, and ensure that everything looks in order:

```{r, comment=""}
# prints contents of "co_counties_race"
co_counties_race
```

As always, it is also possible to view the dataset in the R Studio data viewer by running ```View(co_counties_race)```.

## Clean the *tidycensus* dataset

Having extracted the dataset, you may want to tidy it up, depending on your needs and preferences. For example, the "NAME" field includes the name of the state, which is not really necessary here since there are only observations from Colorado in the dataset. The code below creates a new variable named "County", which is populated by removing the state name from the "NAME" field, and updates the ```co_counties_race``` object with this new variable:

```{r}
# Remove state name from NAME field
co_counties_race<-co_counties_race %>% 
                  mutate(County=str_remove_all(NAME, ", Colorado"))   
```       

```{r, echo=F}
co_counties_race<-co_counties_race %>% relocate(GEOID, County)
```

Note the newly created "County" variable is now in the dataset: 

```{r, comment=""}
# Prints contents of "co_counties_race"
co_counties_race
```

##  Define new variables

Now that we have a cleaned dataset with all our necessary variables, we can use these variables to generate the demographic variables needed to calculate the bias index. First, the code below defines a new variable, called "total_pop_over17", that is calculated by subtracting the total population that is 17 and under from the total overall population:

```{r}
# Create variable for total over-17 population
co_counties_race<-
  co_counties_race %>% 
     mutate(total_pop_over17=
              total_pop-totalpop_men_u5-totalpop_men_5to9-
              totalpop_men_10to14-totalpop_men_15to17-totalpop_women_u5-
              totalpop_women_5to9-totalpop_women_10to14-totalpop_women_15to17)
```

Then, we create a new variable named "total_black_pop_over17", which is defined by subtracting the total Black population that is 17 and under from the total Black population:

```{r}
# Create variable for total over-17 black population
co_counties_race<-
  co_counties_race %>% 
  mutate(total_black_pop_over17=
          black_totalpop-black_men_u1-black_men_1-
          black_men_2-black_men_3-black_men_4-black_men_5-black_men_6-
          black_men_7-black_men_8-black_men_9-black_men_10-black_men_11-
          black_men_12-black_men_13-black_men_14-black_men_15-black_men_16-
          black_men_17-black_women_u1-black_women_1-black_women_2-black_women_3-
          black_women_4-black_women_5-black_women_6-black_women_7-black_women_8-
          black_women_9-black_women_10-black_women_11-black_women_12-
          black_women_13-black_women_14-black_women_15-black_women_16-
          black_women_17)
```

## Finalize and export the dataset

Now that we have our two key variables defined, let's clean up the dataset by removing the variables we no longer need, only keeping the variables necessary to help create the bias index. Below, we take the existing dataset assigned to ```co_counties_race```, and select the "GEOID" and "County" variables (which serve as ID variables), "total_black_pop_over17" and "total_pop_over17" (which are used to compute the bias index), and "total_pop" (which is not necessary to create "bias_index", but which could prove useful in exploring alternate ways of defining a bias index than the one implemented in the tutorial). The new dataset is assigned to an object named ```co_counties_census_2010```:

```{r}
# Clean data by select relevant variables for analysis, and assign 
# selection to new object named "co_counties_census_2010"
co_counties_census_2010<-
  co_counties_race %>% 
    select(GEOID, County, total_pop, total_black_pop_over17, total_pop_over17)
```

Let's now view the contents of ```co_counties_census_2010```:

```{r, comment=""}
# prints contents of "co_counties_census_2010"
co_counties_census_2010
```

At this point, we now have the census data that was provided to you at the start of the tutorial. 

To export this dataset, use the **```write_csv()```** function; below, the first argument is the name of the object which contains the dataset to be exported, and the second argument is the desired file name. The data is exported to the current working directory, and can subsequently be opened on your spreadsheet software of choice as a CSV file. 

```{r, eval=FALSE}
# Exports the data assigned to the "co_counties_census_2010" object
write_csv(co_counties_census_2010, "co_counties_census_2010.csv")
```



